---
title: "Aligning Records"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aligning Records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(ecoeval)
ex <- function(f) system.file("extdata", f, package = "ecoeval")
```

Before any accuracy number means anything, we have to decide which AI row
corresponds to which gold row. That is the core of the app, and it is why the
human stays in the loop: a person looking at two rows side by side can tell they
describe the same interaction even when the text doesn't match. An automated
matcher often can't.

## Scope: papers are a filter, not a scored entity

**Papers: intersection. Records: union.** Scope is the papers present in both
paper sets — each source's paper list when supplied, otherwise the papers
appearing in its records. Within those papers, every record from both sides
appears; some pair up, some don't.

```{r}
schema <- read_schema(ex("schema.json"))
ai   <- prepare_records(read_table_any(ex("ai_records.csv")),   "doi", prefix = "a")
gold <- prepare_records(read_table_any(ex("gold_records.csv")), "doi", prefix = "g")

ai_papers   <- prepare_papers(read_table_any(ex("ai_papers.csv")),   "doi")
gold_papers <- prepare_papers(read_table_any(ex("gold_papers.csv")), "doi")

scope <- compute_scope(paper_set(ai, ai_papers), paper_set(gold, gold_papers))
lengths(scope)
```

Papers outside the intersection are excluded, never penalised. A gold-only paper
simply narrows the comparison set; it carries no warning.

Paper alignment determines scope, not score. Its asymmetry is worth knowing: a
*missed* paper match just shrinks the evaluation set, but a *wrong* paper match
compares one paper's records against another's and produces garbage. That's what
the paper-link review guards against.

## The comparator cascade

Comparison is a ladder, not a single test: **exact → normalized → fuzzy → LLM
judge**. Each rung sees only what the previous couldn't resolve, so the
expensive rungs run on a small minority of cells.

```{r}
compare_pair("predation", "predation", "fuzzy")$rung
compare_pair("Predation ", "predation", "fuzzy")$rung
compare_pair("Myotis lucifugus", "Myotis lucifigus", "fuzzy")$rung
```

A field's configured comparator names the **highest** rung the cascade may climb
to, so it can't quietly accept more than you asked for:

```{r}
compare_pair("Predation", "predation", "exact")$agree
compare_pair("Myotis lucifugus", "Myotis lucifigus", "normalized")$agree
```

Defaults come from the JSON Schema type — the only inference available without
domain knowledge, and enough:

```{r}
schema$fields[, c("field", "type", "comparator")]
```

Without normalisation the first run is a wall of disagreement that is mostly
formatting noise. Two of the defaults exist purely to absorb it:

```{r}
compare_pair("10", "10.0", "numeric")$agree
compare_pair("2021-03-08", "08/03/2021", "date")$agree
```

That second one matters more than it looks. `strptime()` happily ignores
trailing characters, so a naive parser reads `"05/01/2020"` under `"%Y/%m/%d"`
as the year 5. `parse_date_loose()` accepts a format only when the parsed date
round-trips back to the original text.

## Two LLM-backed strategies, for different columns

|                            | Mechanism                              | Cache key    | Best for                   |
| -------------------------- | -------------------------------------- | ------------ | -------------------------- |
| **Normalize-then-compare** | map both values to a canonical form    | per **value** | species names              |
| **LLM judge**              | compare the pair, return verdict + why | per **pair**  | supporting sentences, prose |

The judge **cannot move earlier** in the pipeline. Alignment evaluates every
candidate pair within a paper — 5 AI records against 5 gold records is 25 pairs
times N linkage fields of LLM calls just to build the comparison matrix.
Normalisation is per value: 10 values, 10 calls, then compare cheaply.
Normalisation scales to N; the judge would scale to N².

So the order is:

1. Pick linkage fields and comparators
2. **Normalise the linkage fields** on both sides — per value, cached, a no-op
   when no normalising comparator was assigned
3. Run the matcher on the normalised values
4. Score cells with the cheap rungs; the judge fires per paper, on paired rows
   only, and only on cells the cheaper rungs couldn't resolve

Two consequences: **display originals, compare normalised**, and **the collapse
check runs on normalised values** — `"M. lucifugus"` and `"Myotis lucifugus"`
should collapse together, and running it raw undercounts.

With no API key configured, the judge rung reports its cells as *pending* rather
than pretending to a verdict:

```{r}
res <- compare_pair("bats ate moths", "moths were taken by bats", "judge")
c(agree = res$agree, rung = res$rung, pending = res$pending)
```

Verdicts are cached per pair, because LLM output varies between calls and a
reproducible run needs them frozen rather than re-derived. The cache is written
into `run_config.json`.

## The matcher

fastLink, blocked per paper, **permissive — no similarity floor**. It proposes
pairings even when key identifiers disagree, and the human judges. Dropping a
pair because the identifiers look wrong is precisely the decision the human
should be making with the values in front of them.

There is one wrinkle worth knowing about. fastLink's EM learns which fields
discriminate *from the data it is given* — and a per-paper block holds a handful
of records. On two records it will confidently pair the wrong ones. So ecoeval
fits the model **once over every scoped record** and then applies it per paper:
pooled estimation, blocked application.

```{r}
config  <- default_comparator_config(schema)
linkage <- config$field[config$linkage]
linkage

paper_map <- tibble::tibble(ai_paper = scope$papers, gold_paper = scope$papers)
pairs <- align_records(ai, gold, linkage, paper_map)
table(pairs$kind)
```

fastLink also clusters string distances internally, and that clustering is
randomly initialised — two identical calls could otherwise return different
pairings. ecoeval pins the seed (see `ecoeval_seed()`), restoring the caller's
RNG state afterwards, so a run is reproducible without disturbing anything else
in your session.

**Matching is 1:1.** Leftovers on either side are false positives and false
negatives, because that is what they are. An extra AI record *is* a false
positive; whether it came from a hallucination or a dedup failure is a diagnosis
you make from the collapse warning, not something the metric needs to model.

## The three operations

At the **row** level you fix the pairing — reject a link, or link two orphans.
At the **cell** level you fix the verdict — assert that two values really do mean
the same thing, or really don't.

**Linking matters more than rejecting.** Rejection can only make the alignment
sparser. The pairs a matcher misses are exactly the ones whose identifiers
disagree — and those are the ones a human can recognise as the same entity.
Without manual linking you could only correct in one direction.

All three recompute every metric immediately, and all three persist in
`run_config.json`. A hand-tuned result should say so:

```{r}
correction_count(
  rejected  = data.frame(ai_rid = character(0), gold_rid = character(0)),
  added     = data.frame(ai_rid = "a1", gold_rid = "g1"),
  overrides = data.frame(pair_id = c("p1", "p2"), field = "x", agree = TRUE)
)
```

## When to review by hand

You don't have to check its work row by row. But a batch pass knows which pairs
look wrong, which turns "review all 47 papers or none" into "look at these
three":

```{r}
cells <- score_cells(pairs, ai, gold, config)
suspect_pairs(cells, linkage)
```

Two cheap signals, both strong. A pair that agrees on **nothing** is almost
certainly two unrelated records the matcher joined. A pair that disagrees on
every **identity column** is the subtler case: it may agree on incidentals like
country and year while the fields that establish which record this actually *is*
all disagree.

## Granularity: do the two datasets agree what a record is?

Group the gold records by the chosen linkage fields within each paper and count
collapses. If many rows collapse together, the human distinguished records by
something the key doesn't capture.

```{r}
granularity_check(gold, linkage, "gold")
```

Two gold rows in that paper differ only by year, which isn't in the identity
key. Add a field and try again — or accept that the two datasets are counting
different things, which is itself a finding.
