---
title: "Evaluation Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluation Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

You have two sets of results from the same papers: one the AI produced, one a
person produced by hand. You want to know how well the AI did.

You can't just lay the two tables side by side, because people — and AI — often
describe the same thing in different ways. So ecoeval works in two steps. First
it **pairs up the rows**, matching each AI row to the human row describing the
same thing. Then it **scores the result**.

This guide walks the whole thing. [Aligning
records](aligning-records.html) covers why the pairing is the hard part;
[Metrics](metrics.html) covers how the numbers are computed.

## The five inputs

Four data inputs, symmetric across the two sources, plus the schema:

|          | Paper list       | Record list        |
| -------- | ---------------- | ------------------ |
| **AI**   | papers processed | records extracted  |
| **Gold** | papers reviewed  | records recorded   |

`schema.json` — the schema the AI extracted against — is a **required fifth
input**. It drives comparator defaults, the enum conformance check, and the
default linkage-field suggestion.

The paper lists are optional. They **widen scope**; they don't change how
anything is scored. Each one lets you see a case that was otherwise invisible:

| List supplied | Papers it brings into scope        | What that lets you catch |
| ------------- | ---------------------------------- | ------------------------ |
| Gold          | Human reviewed, found nothing      | AI **over**-extraction   |
| AI            | AI processed, found nothing        | AI **under**-extraction  |

For an ecoextract database the AI paper list is free — it's the `documents`
table, which knows every paper processed regardless of whether extraction
produced records.

## Running it

```r
library(ecoeval)

# Launch and pick inputs in the app.
run_eval_app()

# Or supply them up front.
run_eval_app(
  ai          = "ecoextract_records.db",
  gold        = "gold_standard.csv",
  schema      = "ecoextract/schema.json",
  gold_papers = "papers_reviewed.csv"
)
```

The app is a wizard with a persistent progress rail. Stages are revisitable
without redoing everything downstream:

1. **Load** — the five inputs
2. **Identify the papers** — which column identifies a paper; detected in a
   fixed priority order (DOI, file name, title, first author + year)
3. **Align papers** — sets scope
4. **Map record fields** — column mapping, comparators, identity columns
5. **Compare** — one paper at a time; alignment review *and* results on one screen
6. **Dashboard** — aggregate metrics, updating as review progresses

To try it without any data of your own, press **Use the bundled example** on the
load screen. It reads the synthetic fixtures shipped with the package, each row
of which exists to exercise a specific case the evaluation has to get right.

## Without the app

The app is a shell over a library. Everything it computes is a plain function
over data frames, so an evaluation can be scripted:

```{r}
library(ecoeval)

ex <- function(f) system.file("extdata", f, package = "ecoeval")

run <- evaluate_extraction(
  ai          = ex("ai_records.csv"),
  gold        = ex("gold_records.csv"),
  schema      = ex("schema.json"),
  ai_papers   = ex("ai_papers.csv"),
  gold_papers = ex("gold_papers.csv")
)
```

Scope first — papers are a filter, not a scored entity:

```{r}
str(run$scope)
```

Then the numbers:

```{r}
aggregate_metrics(run$cells)
record_metrics(run$pairs)
```

And the per-column picture, sorted worst first, which is the triage order:

```{r}
field_metrics(run$cells)[, c("field", "n_scored", "accuracy", "precision", "recall")]
```

## Two ways to use this, and neither is the shortcut

**"How accurate is it?"** — press **Resolve all differences** on the dashboard.
The judge runs across every scoped paper, the numbers fill in, and you never
open a single paper. This is the common case.

**"Why is it wrong?"** — walk the papers, read the mismatches, spot the patterns
worth fixing before the next extraction run. This is what you do when you're
iterating on the schema or the prompt rather than reporting a number.

Skipping review means accepting the matcher's pairings. Field-level metrics are
largely robust to that; record-level precision and recall are not. Worth
knowing, not worth worrying about.

The dashboard is honest about which path produced the numbers:

```{r}
progress_summary(run$cells, run$scope$papers)
```

`n_pending_cells` counts cells the cheap comparator rungs couldn't settle and
the judge hasn't seen. They currently count as disagreements.

## Findings

Findings are **output, not gates**, grouped by who acts on them — the schema,
the model, or the gold standard. Nothing here adjusts a metric.

```{r}
run$findings[, c("group", "finding")]
```

Only two conditions actually stop the app, because neither leaves anything to
compare: zero paper overlap between the two sources, and no mappable fields
between the two record sets.

## Export

The session artifact and the export bundle are the same object, so save,
restore, and download share one implementation.

```{r, eval = FALSE}
export_bundle(
  "ecoeval_run_2026-08-31",
  cells  = run$cells,
  pairs  = run$pairs,
  config = capture_run_config(new_run_config(), comparators = run$config),
  scope  = run$scope,
  schema = run$schema
)
```

```
ecoeval_run_2026-08-31/
  aligned_table.csv        one row per aligned record, the values in full
  aligned_table_long.csv   tidy: one row per record x field, for pivoting
  metrics.xlsx             one sheet per field + summary sheet
  findings.csv             findings, grouped schema / model / gold
  schema_patch.json        proposed schema edits
  plots/                   column accuracy, per-field confusion matrices (PNG)
  run_config.json          mappings, comparators, linkage fields, judge verdict
                           cache, every manual link/reject decision, scope
  README.txt               what's here, scope statement, any active warnings
```

`run_config.json` is what makes a run reproducible. Reload it with
`run_eval_app(run_config = "...")` and you are exactly where you were — manual
link decisions and cached LLM verdicts included.

## Iterating

Two things make this an improvement tool rather than a report generator.

**Runs are versioned and diffable.**

```{r}
before <- capture_run_config(new_run_config(),
                             metrics = list(record_f1 = 0.71, interaction_type = 0.62))
after  <- capture_run_config(new_run_config(),
                             metrics = list(record_f1 = 0.78, interaction_type = 0.91))
diff_runs(before, after)
```

**Schema findings emit an applicable patch, not prose.**

```{r}
patch <- schema_patch(run$conformance, NULL, run$cells, run$schema)
patch$enum_additions
```

A value recurring many times is a category the schema is missing. A single one
is a typo. The counts make the difference obvious, and the two have opposite
fixes — so the patch proposes only the first.

## Requirements

- R >= 4.0
- [ecoextract](https://github.com/n8layman/ecoextract) output, or any record
  table plus the `schema.json` it was extracted against
- An API key for the LLM judge — copy `.env.example` to `.env`. Without one the
  cheap comparator rungs still run; contested cells are reported as unjudged
  rather than silently scored wrong.
