---
title: "Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(ecoeval)

run <- evaluate_extraction(
  ai     = system.file("extdata", "ai_records.csv",   package = "ecoeval"),
  gold   = system.file("extdata", "gold_records.csv", package = "ecoeval"),
  schema = system.file("extdata", "schema.json",      package = "ecoeval")
)
```

## One unified accounting

Computed over the **entire set** — every record from both sides, matched or not.
Per column, each cell contributes:

```{r}
state_contributions()
```

Reading those states:

| State                     | Outcome            | Colour | Meaning                                       |
| ------------------------- | ------------------ | ------ | --------------------------------------------- |
| `agree`                   | agree              | green  | matched pair, values agree under the comparator |
| `disagree`                | disagree           | purple | matched pair, values differ                   |
| `ai_missing`              | only in the gold   | yellow | matched pair, the AI left it blank            |
| `gold_only`               | only in the gold   | yellow | gold record with no counterpart               |
| `gold_missing`            | only in the AI     | orange | matched pair, the human left it blank         |
| `ai_only`                 | only in the AI     | orange | AI record with no counterpart                 |
| `blank`                   | agree              | green  | neither side has a value                      |

Seven states, four colours. Whether a one-sided value came from an unpaired
record or from a blank on one side of a paired one is a detail of the
*alignment*, not of the finding, so `cell_outcome()` collapses them; and a
column neither side filled in is agreement about absence. The per-paper
heatmap and the exported table paint the result the same way. The overview on
the dashboard cannot: its tiles cover several records at once, so it shades the
share that agree rather than claiming one of these four.

The colours are not the accounting: a `blank` cell is a **true negative** and
drops out of accuracy, precision and recall entirely, as the table above shows.
It reads as agreement and counts as nothing.

A disagreement counts as **both** a false positive and a false negative. That's
the standard multi-class treatment: a misclassification is a false positive for
the value asserted and a false negative for the value missed.

Cells blank on both sides are true negatives and drop out of precision and
recall, because a field neither party filled in says nothing about extraction
quality.

## The colours are the accounting

Every cell is one of five things, and the dashboard's four colours are exactly
those boxes, so counting the colours *is* computing the confusion matrix.

```{r}
ct <- confusion_totals(run$cells)
ct$by_outcome[, c("outcome", "colour", "n", "tp", "fp", "fn", "tn")]
```

Which ties out with the per-field accounting by construction — both are
`state_contributions()` summed, so there is no second calculation to drift:

```{r}
fm <- field_metrics(run$cells)
c(tp = ct$tp - sum(fm$tp), fp = ct$fp - sum(fm$fp), fn = ct$fn - sum(fm$fn))
c(accuracy = ct$accuracy, precision = ct$precision, recall = ct$recall, f1 = ct$f1)
```

## The property the accounting rests on

A wrongly linked pair that disagrees on everything produces FP + FN per
column — **exactly the same as leaving those rows unlinked**, where the AI-only
row gives FP and the gold-only row gives FN.

```{r}
ai   <- tibble::tibble(.rid = "a1", .paper = "P", v = "alpha", w = "one")
gold <- tibble::tibble(.rid = "g1", .paper = "P", v = "omega", w = "two")
config <- default_comparator_config(fields = c("v", "w"))

linked <- score_cells(
  tibble::tibble(pair_id = "p1", paper = "P", ai_rid = "a1", gold_rid = "g1"),
  ai, gold, config)

unlinked <- score_cells(
  tibble::tibble(pair_id = c("p1", "p2"), paper = "P",
                 ai_rid = c("a1", NA), gold_rid = c(NA, "g1")),
  ai, gold, config)

identical(
  field_metrics(linked)[, c("field", "tp", "fp", "fn")],
  field_metrics(unlinked)[, c("field", "tp", "fp", "fn")]
)
```

So linking earns nothing except where fields genuinely agree, and bad links
cannot inflate field accuracy. **Field-level metrics are largely robust to
alignment error.** Record-level metrics are not — pairing two records converts
1 FP + 1 FN into 1 TP — which is exactly the judgment the human review exists
for.

## Two confusion matrices, both about extraction

Plainly labelled. There is no third matrix about linkage quality; users
shouldn't have to learn that distinction.

**Records** — did the AI find the right rows?

```{r}
record_metrics(run$pairs)
```

**Fields** — did it fill them in correctly?

```{r}
field_metrics(run$cells)[, c("field", "tp", "fp", "fn", "n_scored", "accuracy")]
```

Sorted worst accuracy first, which is the triage order. But read it carefully: a
column that is uniformly wrong has three possible causes and the tool can't tell
them apart.

1. **A misconfigured comparator** — exact matching on a date column with two
   formats, or a fuzzy threshold set too tight. On a first run this is the most
   likely explanation, and the cheapest to check.
2. **An ambiguous field description** — the prompt didn't pin down what to put
   there.
3. **The model genuinely failing** on that field.

Check them in that order. The chart tells you *where* to look, not why.

## The two aggregates

Two numbers, plain labels — no micro/macro jargon.

```{r}
aggregate_metrics(run$cells)
```

`overall_accuracy` weighs every scored cell the same; `column_mean_accuracy`
weighs every column the same. They diverge when fill rates are uneven, and the
gap is itself informative.

## Column-level matrices

The form adapts to the column's type, because a K×K matrix over thousands of
species names is useless.

**Enum, boolean, or low cardinality** → a true K×K, gold class by AI class, with
an explicit `(not in schema)` row and column so out-of-enum values appear rather
than being silently dropped. Those are among the most interesting cells on the
chart.

```{r}
cc <- column_confusion(run$cells, "interaction_type", run$schema)
cc$type
subset(cc$matrix, n > 0)
```

**Free text or high cardinality** → 2×2 on presence and absence, with the
populated-by-both cell **split into "correct" and "wrong value"**. That split is
what makes it useful: it separates *didn't populate the field* from *populated
it wrong*, which have entirely different causes.

```{r}
cc <- column_confusion(run$cells, "all_supporting_source_sentences", run$schema)
cc$type
cc$matrix
```

**Numeric or date** → presence and absence plus an error distribution.

```{r}
cc <- column_confusion(run$cells, "sample_size", run$schema)
cc$errors
```

Each result carries `notes` — cautions such as one class dominating (accuracy is
then trivially high and uninformative) or too few matched records to mean
anything.

## Fill rates

Per-column fill rates for both sources, side by side. A field humans always
populate and the AI rarely does is usually a field-description problem, not a
model failure — which is why it lands in the *schema* group of the findings, not
the *model* group.

```{r}
fill_rates(run$cells)
```

## Schema conformance is not folded into anything

Validate both sides against the schema, warn, mark the offending cells, and stop
there. Scoring treats both sides at face value.

```{r}
subset(run$conformance, source == "gold")
```

- **AI side** — an invalid value is simply a **true error**, scored as wrong like
  any other. Structured output should have prevented it, so it's also worth
  flagging as a pipeline problem, but it needs no special accounting.
- **Gold side** — you go and fix your gold standard. Not something the tool
  models.

No ceiling arithmetic, no reclassification, no confirmation flow. The counts do
the work: `commensalism` appearing three times is a category the schema is
missing; `predatoin` appearing once is a typo. The two have opposite fixes.

Schema violations are **marked, not coloured**. Validity is
orthogonal to match state, a cell can be both, and either side's value (or both)
may be the offender. Fill carries agreement; the marker carries validity.

## Honest reporting

The dashboard distinguishes three progress states, because they mean different
things:

| State        | Meaning                                                  |
| ------------ | -------------------------------------------------------- |
| **Unjudged** | cheap comparator rungs only; contested cells unresolved  |
| **Judged**   | the LLM settled the contested cells                      |
| **Reviewed** | a person looked at it                                    |

```{r}
progress_summary(run$cells, run$scope$papers)
```

*"47 of 47 papers judged · 3 reviewed"* is an honest description of a fast-path
result, and it tells a later reader exactly how the numbers were produced
without editorialising about it. It travels with the export, in `README.txt`,
alongside the scope statement and any active warnings — so the caveats don't
live only on screen.
